---
title: "VA with CppAD for GSM"
author: Benjamin Christoffersen
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  error = FALSE, cache = "./README-cache/", fig.path = "figures/README-", 
  echo = TRUE)
options(digits = 4, scipen = 7)
.fig_height_small <- 4
palette(c("#000000", "#009E73", "#e79f00", "#9ad0f3", "#0072B2", "#D55E00", 
          "#CC79A7", "#F0E442"))
```

This package is made to 

 - show implementations of Laplace approximations and Gaussian quadrature 
   using `CppAD` through `TMB`. 
 - show implementations of variational approximation in generalized survival 
   models (GSM).
 - how to use `TMB` to get extra operators for `CppAD` and make `CppAD` work 
   with `Eigen`. 
 - use `TMB` and `Rcpp` together and have multiple C++ functions which can  
   be used for R without one function with a switch statement.
   
The latter may be advantageous later as we can work more solely in C++. You 
can contrast the solution here with the discussion made in 
[this post](https://github.com/kaskr/adcomp/issues/247#issuecomment-475002639).
We start by using a temporary directory as our working directory throughout
the rest of the code

```{r set_dir}
old_dir <- getwd()
dir_use <- tempdir()
knitr::opts_knit$set(root.dir = dir_use)
```

We also load a few libraries

```{r load_lib}
library(GSMVAEx)
library(TMB)
```

## log-log GSM
So far, we just compute the gradient of log-log GSM in `C++` and compare it 
with using `TMB` from R. Here is the `loglogTMBEx.cpp` file which we will use
in `R`

```{r show_loglog_code, engine="C++", code = readLines("README/loglogTMBEx.cpp"), eval = FALSE}
```

We make a copy of the file in the temporary directory and compile the code.

```{r cmp_loglog_code, cache = 1}
file.copy(file.path(old_dir, "README", "loglogTMBEx.cpp"),
          "loglogTMBEx.cpp")
stopifnot(compile("loglogTMBEx.cpp") == 0)
dyn.load(dynlib("loglogTMBEx"))
```

Then we have a bit of R code which we need in order to simulate the data
and which we use in subsequent code (this is some code I received and there 
seems to be a bug in it)

```{r tmb_loglog_from_R, cache = 1, dependson = "cmp_loglog_code"}
require(splines)
set.seed(12345)
x <- seq(0, 1, length=1e4)
time <- rweibull(length(x), 1, exp(.1+0.1*x))
censor <- runif(length(x), 0, 5)
event <- time < censor
tobs <- pmin(censor, time)
formula <- ~ x + ns(log(tobs) , df = 1)

getXD <- function(formula, data, newdata, time, eps = 1e-5) {
    newdata[[time]] <- newdata[[time]] + eps
    upper <- rstpm2:::predict.formula(formula, data, newdata)
    newdata[[time]] <- newdata[[time]]-2*eps
    (upper-rstpm2:::predict.formula(formula,data,newdata))/2/eps
}

# TODO: something seems to go wrong here? At least the dimension of beta 
# which we assign later is incorrect in the original code
X <- rstpm2:::predict.formula(
  formula, data.frame(tobs = tobs[event], x = x[event]), data.frame(tobs, x))
head(X) # why is there no `ns(log(tobs) , df = 1)` column?

XD <- getXD(formula, data.frame(tobs = tobs[event], x = x[event]), 
            data.frame(tobs, x), time = "tobs")

MakeADFun <- function(...) {
  newobj <- TMB::MakeADFun(...)
  newobj$call <- match.call()
  newobj
}
```

Finally, we compare the computation of the gradient in C++ in this package 
with the one using the `TMB` interface from `R`

```{r comp_grads_loglog, cache = 1, dependson = "tmb_loglog_from_R"}
local({
  # assign functions to do the recording and compute the gradient
  tmb <- function(){
    f = MakeADFun(
      data = list(X = X, XD = XD, tobs = tobs, event = as.double(event),
                  eps = 1e-6, kappa = 1.0),
      # TODO: original code
      # parameters = list(beta = c(-5, 0, 1)), method = "nlminb",
        parameters = list(beta = c(-5, 0   )), method = "nlminb",
      DLL = "loglogTMBEx", silent = TRUE)
    f$gr(c(-5, 0))
  }
  imp <- function()
    GSMVAEx::loglog(tobs = tobs, event = as.double(event), XD = XD, X = X, 
                    eps = 1e-6, kappa = 1.0, beta = c(-5, 0))
  
  # we get the same gradient
  stopifnot(all.equal(drop(tmb()), drop(imp())))
  
  # the speed is the same
  microbenchmark::microbenchmark(TBM = tmb(), Implementation = imp(), 
                                 times = 100)
})
```

Of course, one of the nice things about `MakeADFun` is that we do not have 
to do the recording over and over again and save a lot of computation time. 
We can similarly save a pointer to a C++ object in our implementation. 
Finally, to make the point 
that we can call multiple different C++ functions from R 
then we make a call to the
`hello_world_func` function in `src/R-interface.cpp` file shown below

```{r show_r_interface, engine="C++", code = readLines("src/R-interface.cpp"), eval = FALSE}
```

here

```{r hello_world}
hello_world_func()
```

